---
title: "Introduction to R, Data types and Data Structure in R"
author: :"Siddharth Chaudhary"
subtitle: "Data in R, Fall 2022"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# --------------
## Getting help
# --------------


```{r}
help(sqrt)
?sqrt
help.start()
```

# --------------
## basic operations
# --------------

```{r}
2+2
2-3*4
(2-3)*4
5/9*4
3^2   # 3 to the power of 2

5 < 4

3 == 3   # this is a test, are these values equal?

6/3 != 2  # this is a test, are these values NOT equal

8/2 <= 4  # this is a test, is the left side value less than or equal to the right side?
```

# --------------
## Basic functions
# --------------

```{r}
sqrt(4)   #square root of four
sqrt(9)   #notice that comments after the pound sign are ignored

# notice that case matters in R
sqrt(9)

print("Help! I've been abducted by aliens!!!")

cat("Help!\n\n", "I've been abducted by smart aliens!!!", sep="")

```

Notice that the output dumps onto the console. Whatever is dumped to the console is not retained.

We can change that by assigning the results to a object.

```{r}
my_results <- sqrt(9)
my_results
#My_results   # case matters

my_results = sqrt(16)
my_results

# Better practice is to use assignment operator
```

# --------------
## command separation
# --------------

```{r}
print("blah")
print("ughh")
print("really?")

print("blah"); print("ughh"); print("really?")  # semi colons for separation
# we can also string a command across multiple lines, though this is not good practice

# notice that R adds a + sign to the new line to indicate continuation of the command
print("blah"
)
```

In this lecture, you will learn about the most basic data structure in R.
# --------------
# Part 1 - Basic data types
# --------------

There are four common types of data in R: double (numeric), integer, logical and character. (We won't work with raw or complex values.)

# --------------
## Numeric
# --------------

**double** or **numeric** vector
```{r double vector typeof}
vect_dbl <- c(1.1,-2.5,4,6,2.2,12) # create vector data object
vect_dbl
typeof(vect_dbl)
```

We can also examine a data object using the **str** structure function.
```{r str command}
str(vect_dbl)
```

Notice that the **str** function tells us this is numeric data!

# --------------
## integer
# --------------

**integer** vector
```{r integer vector}
vect_int <- -1:10 # create vector data object
vect_int
typeof(vect_int)
str(vect_int)
```

R treats integers the same as numeric values for most functions.

# --------------
## logical
# --------------

**logical** vector
```{r logical vector}
vect_log <- c(T, FALSE, TRUE, F,T,T) # create vector data object
vect_log
typeof(vect_log)
str(vect_log)
cat("\nsum of logical vector: ", sum(vect_log))
```

Note that R treats T and TRUE, F and FALSE as equivalent.
When treated as numbers, T=1, F=0.

**logical** and numeric vector
```{r logical and numeric vector}
vect_log <- c(T, FALSE, TRUE, F,T,T, 5.5) # create vector data object
vect_log
typeof(vect_log)
str(vect_log)
cat("\nsum of logical vector: ", sum(vect_log))
```

# --------------
## character
# --------------

**character** vector
```{r character vector}
vect_chr <- c("Dave", "Vlad","Chuck","Larry","Bob","Mary","Doug","Daryll","Ron","Sue", "Laura", "Rich")
vect_chr
typeof(vect_chr)
str(vect_chr)
```

Note that each new line in the vector printout is labeled with the vector position. Character values are represented with quotes. 

The structure function will list specific character values as space allows, then...  

**character and numeric** vector
```{r character and numeric vector}
vect_chr <- c("Dave", "Vlad","Chuck","Larry","Bob","Mary","Doug","Daryll","Ron","Sue", "Laura", "Rich", 5.5)
vect_chr
typeof(vect_chr)
str(vect_chr)
```

# --------------
## Atomic vector
# --------------

An **atomic vector** or **vector** in R is a one-dimensional data structure. A **vector** holds a string of values, all of the same type, with the string having (practically) any length. You can create a vector using the combine or concatenate **c** function.
```{r vector 1}
vect <- c(1,-2,4,6,2.2,12) # create vector data object
vect  # print data object to the console
```

Notice how the vector is printed to the console.

You can test whether an object is an atomic vector using two different commands.
```{r vector test}
is.atomic(vect)   # more robust test
is.vector(vect)   # won't always work, depending upon attributes
```

# --------------
## Vector properties
# --------------

We can assign and extract vector properties.

We can obtain the vector length with the following command.
```{r vector length}
# length function gives length of a vector (number of values stored)
length(vect)
length(c("123456789","0"))
```

There are six pieces of data, six positions in this vector.

We can name elements in a vector.
```{r vector name example}
# Create a vector with my height and weight in SI units (m, kg)
vect_example <- c(1.75, 70)
vect_example

cat("\n")   # this prints a line space

#now assign names for meaning, notice these are character strings
names(vect_example) <- c("height (m)", "weight (kg)")
vect_example

cat("\n")   # this prints a line space
names(vect_example) <- c("height (m)", "weight (kg)")
vect_example
```

We can also assign vector names when creating the vector.
```{r vector name}
# Create a vector with my height and weight in SI units (m, kg)
vect_example <- c("height (m)"=1.75, "weight(kg)" =70)
vect_example
```

We can inquire about vector type using **typeof()** function.
```{r vector type}
typeof(vect_example)
typeof(c("123456789","0"))
```

"double" means double precision numeric or commonly thought of as "numeric". We will discuss other types of data.

Finally we can inquire about data object attributes.
```{r vector attributes}
vect <- c(1,-2,4,6,2.2,12) # create vector data object
attributes(vect)
cat("\n")
attributes(vect_example)
```

The first vector has no attributes. The second has names. In object oriented programming, data object attributes can be considered "meta data".

# --------------
## Creating Vectors
# --------------

You can also create vectors with other commands.
```{r vector create functions}
vect_rep <- rep(1.2,times=10) # replicate value 10 times
vect_rep

vect_seq <- seq(from=1, to=13, by=2) # sequence from 1 to 12 by 2
vect_seq

vect_seq_by1 <- 1:6 # sequence integers by 1
vect_seq_by1

vect_seq_by1_neg <- -1:6 # sequence integers by 1
vect_seq_by1_neg

vect_seq_by1_pr <- -(1:6) # sequence integers by 1
vect_seq_by1_pr
```

# --------------
## Missing values (NA)
# --------------

All spaces in a vector or matrix must be filled with something. If there is no data for a position, that position is represented by the missing value designation in R, **NA**. Note that a missing value is NOT the same as the character value "NA".

Replace Doug with NA and Sue with "NA".
```{r character NA vector}
vect_na <- c("Dave", "Vlad","Chuck","Larry","Bob","Mary",NA,"Daryll","Ron","NA", "Laura", "Rich")
vect_na
str(vect_na)
```

```{r all NA vector}
vect_rep_na <- rep(NA,85)
vect_rep_na
str(vect_rep_na)
```
# --------------
# coerce data types
# --------------
We can *coerce* data into a particular type or from one type to another.

## as.character

**as.character** examples
```{r as.character}
vect <- c(1.1,-2.5,4,6,2.2,12) # create vector data object
cat("numeric to character \n")  # cat prints nicely
vect
as.character(vect)

# we put a \n in front for an extra return or line
vect <- -1:8 # create vector data object
cat("\ninteger to character\n")
vect
as.character(vect)
cat("\n")

vect <- c(T, FALSE, TRUE, F,T,T) # create vector data object
cat("\nlogical to character\n")
vect
as.character(vect)
```

## as.numeric

**as.numeric** examples
```{r as.numeric}
cat("integer to numeric\n")
vect
as.numeric(vect)
str(as.numeric(vect))

cat("\nlogical to numeric\n")
vect
as.numeric(vect)

vect <- c("Dave", "Vlad","Chuck","Larry","Bob","Mary","Doug","Daryll","Ron","Sue", "Laura", "Rich")
cat("\ncharacter to numeric\n")
vect
as.numeric(vect)
```

Oops! Some data can't be coerced!

## as.integer

**as.integer** examples
```{r as.integer}
cat("numeric to integer\n")
vect2 <- c(1.1,-2.6,4,6.6,2.2,12)
vect2
as.integer(vect2)
```

*as.integer* rounds off numeric values, otherwise has effect like **as.numeric**.

## as.logical

**as.logical** examples
```{r as.logical}
cat("numeric to logical\n")
vect1 <- c(1.1,-2.6,0,6.6,2.2,12)
vect1
as.logical(vect1)

cat("\ninteger to logical\n")
vect2
as.logical(vect2)

cat("\ncharacter to logical\n")
vect4 <- c("Dave", "Vlad","Chuck","Larry","Bob","Mary","Doug","Daryll","Ron","Sue", "Laura", "TRUE")
vect4
as.logical(vect4)
```

## coercion of mixed data

Finally, data are also coerced when creating a vector, if necessary, since vectors can only contain one type of data.
```{r coercion vector creation}
vect5 <- c(T, F, 1, T)
str(vect5)

cat("\n")
vect6 <- c(T, F, "bob", T)
str(vect6)

cat("\n")
vect7 <- c(1, 2.3, "bob",7)
str(vect7)

```

Logical is coerced to numeric.
Everything is coerced to character if there is a character element.
As we have seen previously, when you combine numeric and integers, a numeric vector results.

**Notes**

* Where coersion fails, NA values result.
* MANY functions coerse data and objects into what is needed. This can be both useful and dangerous!

# --------------
# Matrices
# --------------

A **matrix** in R is a 2-dimensional data structure. A **matrix** holds a n by m grid of values, with each of the two dimensions (practically) any length. As with atomic vectors, a matrix must contain values of just one type.

# --------------
## Creating a matrix
# --------------

You can create a **matrix** using the  **matrix** function. 
```{r matrix create}
# 'data' is a vector
# 'nrow' gives the number of rows
# 'ncol' gives the number of columns
# 'byrow' by default is FALSE, so data is read in column-wise
vect <- c(1,-2,4,6,2.2,12)
mat <- matrix(data = vect, nrow = 3, ncol = 2)
mat
```

## Examine matrix properties

We can examine the properties of this matrix data object.
```{r matrix properties}
# dim function gives the dimensions of a matrix
cat("matrix row-column dimensions:", dim(mat),"\n")

# nrow gives number of rows
cat("\nmatrix row dimension:", nrow(mat),"\n")

# ncol gives number of rows
cat("\nmatrix column dimension:", ncol(mat),"\n")

# typeof gives use the type of data
cat("\nmatrix data type:", typeof(mat),"\n")

# str is also useful
cat("\nmatrix structure:\n")
str(mat)
```

Note that you use *dim* for matrices, *length* to extract the dimensions of vectors.

## name attributes for matrix

We can also name rows or columns by assignment.
```{r matrix }
#first look at attributes
attributes(mat)
cat("\n")    # blank line

#colnames and rownames functions are used to address column and row name attributes
colnames(mat) <- c("Var1","Var2")
mat
cat("\n")
rownames(mat) <- c("a","b","c")
mat

#now look at attributes again
cat("\n")    # blank line
attributes(mat)
```

# use **seq** to generate matrix data

Now we can create a larger matrix with sequenced numbers.
```{r matrix sequence}
# create a larger matrix using sequenced numbers
mat <- matrix(data = 1:20, nrow = 4, ncol = 5)
mat

cat("\n")
str(mat)
```

## Read by rows

What happens if we read in by rows?
```{r matrix 3}
# create a larger matrix using sequenced numbers, read in by row
mat <- matrix(data = 1:20, nrow = 4, ncol = 5, byrow=T)
mat

cat("\n")
str(mat)
```

Note that matrix data is always stored column-wise, even if read in row-wise.

## A few matrix operations

There are many useful matrix operations.
```{r matrix ops}
mat

cat("\nTranspose matrix\n")
t(mat)    # transpose matrix

# num [1:3, 1:2] 1 -2 4 6 2.2 12

cat("\nAnother way to reshape matrix\n")
dim(mat) <- c(5,4)
mat

#cat("\nAnother way to reshape matrix\n")
#dim(mat1) <- c(4,3)
#mat1

cat("\nCreate square matrix\n")
mat <- matrix(c(1,3,5,2,0,0,-1,6,0.5),nrow=3,ncol=3)
mat

cat("\nInvert square matrix\n")
solve(mat) 

cat("\nmultiply matrices\n")
round(solve(mat) %*% mat, 3)  # round off multiplication
```


## functions

For expediency, you can write your own functions

```{r}
?sq   # there is no sq function, but we can create one

sq <- function(x) {
  y <- x*x
  return(y)
}

x <- 4
sq(4)
```

Notice that there is no y or x value in our environment. Variables created within a function are not saved outside of the function.
You need to return these values or otherwise save them if needed. There is a new sq object in our environment.

```{r}
#source("Rlib.txt")
```
