---
title: "Overview: Introduction to R and Data types in R"
author: :"Siddharth Chaudhary"
subtitle: "Data in R, Fall 2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Getting help with R

R includes extensive facilities for accessing documentation and searching for help. The **help()** function and **?** help operator in R provide access to the documentation pages for R functions, data sets, and other objects. 

You can also use the help() function to access information about a package in your library — for example, **help(package="MASS")** — which displays an index of available help pages for the package along with some other information.

Help pages for functions usually include a section with executable examples illustrating how the functions work. You can execute these examples in the current R session via the **example()** command.

```{r}
help.start()   #hypertext based version of R's online documentation
help(sqrt)
?sqrt
example(sqrt)
```

The above methods can be used to get help within R. You can also get help on the internet from sites that specialize for R searches including search.r-project.org and Rseek.org, FAQ listings such as Main R FAQ, Google search engine, Stack overflow and R email lists.

# Basic Operations in R

The syntax for arithmetic operations in R is pretty straight forward as shown below:

```{r}
2+2   # Addition
2-2   # Subtraction
2*2   # Multiplication
5/2   # Division
4**2  # 4 to the power of 2
# some expressions are tests that return Boolean values
2 == 2   
6/3 != 2  
6/3 < 2**2  
# operators precedence and associativity
5+2*3 
(5+2)*3
# logical operations
TRUE & FALSE
```

# Functions in R

R has many built-in functions, including numeric, character, and statistical functions. R also provides the user the ability to add functions. These are called user defined functions. 

For example, R has a built-in function **sqrt()** to find the square root of a number. We will create a function to do the same.
```{r}
sqrt(16)  # built-in function

# user-defined function
sqroot <- function(x) {
  y <- x**(1/2)
  return(y)
}
x <- 16
sqroot(x)
```

# Vectors
An **atomic vector** or **vector** in R is a one-dimensional data structure. A **vector** holds a string of values, all of the same type, with the string having (practically) any length. Atomic vector could be integer, double, character, logical or complex. You can create a vector using the combine or concatenate **c** function.

```{r creating vector}
# double vector
vect1 <- c(1, 4.5, 7, 0, 5.5, -9) 
vect1   # print data object to the console
# numeric vector
vect2 <- c(1, 4, 7)
# logical vector
vect3 <- c(T, FALSE, TRUE, F, T, T, 1, 0)  
# Note: R treats T, TRUE and 1 as equivalent. Similarly with F, FALSE and 0.
# character vector
vect4 <- c("Red", "Blue", "yellow", "violet", 5)  # R converts the 5 into a string
```

There are additional ways to create a vector as shown below:

```{r}
vect2a <- rep(1.2,times=10) # replicate value 10 times
vect2a

vect2b <- seq(from=1, to=13, by=2) # sequence from 1 to 13 by 2
vect2b

vect2c <- -1:5 # sequence integers by 1
vect2c

# We can also assign names while creating or after creating vectors
vect2d <- c("height(m)"=1.75, "weight(kg)" =70)
vect2d
```

## Vector properties

Let's look at some vector properties: type of vector and length of vector.
We can check the types of vector using different functions such as **typeof**, **str** and **class()**. We can obtain vector length with the **length()** command

```{r vector type}
typeof(vect1)
str(vect1)     # this function tells us that it is numeric data
length(vect1)
```

## A few matrix operations

```{r vector operations}
# add two vectors
vect2b + vect2c
# multiply vector by a constant
vect1 * 2
```

# Coerce data types

We can *coerce* data into a particular type or from one type to another. R will automatically convert between built-in object types when appropriate as seen above or we can explicitly convert the data type. We can use the functions **as.character()**, **as.numeric()**, **as.integer()** and **as.logical()**.

```{r}
# numeric to character
vect1
as.character(vect1)
# logical to character
vect3
as.character(vect3)
# logical to numeric
vect3
as.numeric(vect3)
# numeric to integer
vect1
as.integer(vect1)
# numeric to logical
vect1
as.logical(vect1)
```

Some data can't be coerced as shown below.

```{r}
# character to numeric
vect4
as.numeric(vect4)
```

# Matrices

A **matrix** in R is a 2-dimensional data structure. A matrix holds a n by m grid of values, with each of the two dimensions (practically) any length. As with atomic vectors, a matrix must contain values of just one type.

## Creating a matrix

A matrix can be created using the **matrix()** function. 
```{r matrix create}
# 'data' is a vector
# 'nrow' gives the number of rows
# 'ncol' gives the number of columns
# 'byrow' by default is FALSE, so data is read in column-wise
vect1 <- c(1,-2,4,6,2.2,12)
mat1 <- matrix(data = vect1, nrow = 3, ncol = 2)
mat1

# 'data' can also be specified as a sequence of numbers
mat2 <- matrix(data = 1:6, nrow = 2, ncol = 3)
mat2
```

## Matrix properties
```{r matrix properties}
# dim function gives the dimensions of a matrix
dim(mat1)

# nrow gives number of rows
nrow(mat1)

# ncol gives number of rows
ncol(mat1)

# typeof gives use the type of data
typeof(mat1)

# str is also useful
str(mat1)
```

## A few matrix operations

There are many useful matrix operations.
```{r matrix ops}
mat1

# Transpose matrix
t(mat1) 

# Another way to reshape matrix
dim(mat1) <- c(2,3)
mat1

# Create square matrix
mat5 <- matrix(c(1,3,5,2,0,0,-1,6,0.5),nrow=3,ncol=3)
mat5

#Invert square matrix
solve(mat5) 

# multiply matrices
round(solve(mat5) %*% mat5, 3)  # round off multiplication

# add matrices
mat1 + mat2

# sum of rows in matrix mat1
rowSums(mat1)
```